---
title: "Reproducible Research"
author: "Victoria Meo"
date: "July 16, 2016"
output: html_document
---
LOAD AND PREPROCESSING TEH DATA
1.Load the data

```{r echo = TRUE}

{if (!file.exists("Coursera")){
        dir.create("Coursera")}
}
        
{fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile ="C:/Users/Vicky/Desktop/Coursera/activity.zip")
date.Download <- date()
}

# Unzip the file
{unzip(zipfile="C:/Users/Vicky/Desktop/Coursera/activity.zip", exdir="C:/Users/Vicky/Desktop/Coursera/activity")}

# Change working directory to folder 

{setwd("C:/Users/Vicky/Desktop/Coursera/activity")}
activity_raw <- read.csv("activity.csv", stringsAsFactors=FALSE)
```

2. Process/transform the data (if necessary) into a format suitable for your analys.

```{r echo = TRUE}
activity_raw$date <- as.POSIXct(activity_raw$date, format="%Y-%m-%d")

# Compute the weekdays from the date attribute
activity_raw <- data.frame(date=activity_raw$date, 
                           weekday=tolower(weekdays(activity_raw$date)), 
                           steps=activity_raw$steps, 
                           interval=activity_raw$interval)

# Compute the day type (weekend or weekday)
activity_raw <- cbind(activity_raw, 
                      daytype=ifelse(activity_raw$weekday == "saturday" | 
                                     activity_raw$weekday == "sunday", "weekend", 
                                     "weekday"))

# Create the final data.frame
activity <- data.frame(date=activity_raw$date, 
                       weekday=activity_raw$weekday, 
                       daytype=activity_raw$daytype, 
                       interval=activity_raw$interval,
                       steps=activity_raw$steps)

# Clear the workspace
rm(activity_raw)
# Sample of first few rows of the data set
head(activity)
```
WHAT IS THE AVERAGE DAILY ACTIVITY PATTERN

1. Calculate the total number of steps taken per day.
```{r echo = TRUE}
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=TRUE)


names(sum_data) <- c("date", "total")

head(sum_data)
# 2. Make a histogram of the total number of steps taken each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 20), 
     main="Histogram of the total number of steps taken each day\n(NA removed)")
#3.Calculate and report the mean and median of the total number of steps taken per day
mean(sum_data$total)
median(sum_data$total)
```

The mean equals `r mean(sum_data$total)` and the median equals `r median(sum_data$total)`

What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
rm(sum_data)

# Compute the means of steps accross all days for each interval
mean_data <- aggregate(activity$steps, 
                       by=list(activity$interval), 
                       FUN=mean, 
                       na.rm=TRUE)

# Rename the attributes
names(mean_data) <- c("interval", "mean")

head(mean_data)

# Compute the time series plot
plot(mean_data$interval, 
     mean_data$mean, 
     type="l", 
     col="blue", 
     lwd=2, 
     xlab="Interval [minutes]", 
     ylab="Average number of steps", 
     main="Time-series of the average number of steps per intervals\n(NA removed)")
```


2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


```{r}
max_pos <- which(mean_data$mean == max(mean_data$mean))

# We lookup the value of interval at this position
max_interval <- mean_data[max_pos, 1]

max_interval


```

The 5-minute interval that contains the most steps is `r max_interval`


INPUTING THE MISSING VALUES

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r echo = TRUE}

NA_count <- sum(is.na(activity$steps))

```

the number of NA's count equals `r NA_count`

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r echo= TRUE}

na_pos <- which(is.na(activity$steps))

# REPLACES MISSING VALUES WITH THE MEAN STEPS 
mean_vec <- rep(mean(activity$steps, na.rm=TRUE), times=length(na_pos))
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r echo=TRUE}
# Replace the NAs by the means
activity[na_pos, "steps"] <- mean_vec

```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r echo = TRUE}

sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum)

# Rename the attributes
names(sum_data) <- c("date", "total")

# Compute the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day\n(NA replaced by mean value)")

mean(sum_data$total)
median(sum_data$total)
```

The mean is now equal to `r mean(sum_data$total)` and the median is now equal to `r median(sum_data$total)`. By filling in the missing values the mean is increased and so has to median. The mean now equals the median.


ARE THERE DIFFERENCES IN ACTIVITY PATTERNS BETWEEN WEEKDAYS AND WEEKENDS?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r echo=TRUE}
head(activity)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r echo=TRUE}
rm(sum_data)

# Load the lattice graphical library
library(lattice)

# Compute the average number of steps taken, averaged across all daytype variable
mean_data <- aggregate(activity$steps, 
                       by=list(activity$daytype, 
                               activity$weekday, activity$interval), mean)

# Rename the attributes
names(mean_data) <- c("daytype", "weekday", "interval", "mean")

head(mean_data)

xyplot(mean ~ interval | daytype, mean_data, 
       type="l", 
       lwd=1, 
       xlab="Interval", 
       ylab="Number of steps", 
       layout=c(1,2))
```


